---
title: "Reproducible Research: Peer Assessment 1"
author: "Revathi S"
date: "2/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

```{r, echo=TRUE}
## Loading and preprocessing the data
library(dplyr)
library(lubridate)
library(ggplot2)
library(httpuv)

activity <- read.csv("activity.csv", header=TRUE, na.strings = "NA")

# Clean up date class
activity$date <- ymd(activity$date)

# Remove NA
activity1 <- na.omit(activity)
```

## What is mean total number of steps taken per day?
```{r, echo=TRUE}
# Summarize data for ggplot
activity2 <- summarize(group_by(activity1,date),daily.step=sum(steps))
mean.activity <- as.integer(mean(activity2$daily.step))
median.activity <- as.integer(median(activity2$daily.step))

# Plot histogram
plot.steps.day <- ggplot(activity2, aes(x=daily.step)) + 
  geom_histogram(binwidth = 1000, aes(y=..count.., fill=..count..)) + 
  geom_vline(xintercept=mean.activity, colour="red", linetype="dashed", size=1) +
  geom_vline(xintercept=median.activity, colour="green" , linetype="dotted", size=1) +
  labs(title="Histogram of Number of Steps taken each day", y="Frequency", x="Daily Steps") 
plot.steps.day
```

Mean total number of steps taken per day is `r mean.activity`.  
Median total number of steps taken per day is `r median.activity`.

## What is the average daily activity pattern?
```{r,echo=TRUE}
# Prepare data for ggplot
activity3 <- activity1 %>% group_by(interval) %>% summarize(mean.step=mean(steps))

# Plot average number of steps by 5-minute interval
plot.step.interval <- ggplot(activity3, aes(x=interval,y=mean.step)) + 
  geom_line(color="red") + 
  labs(title="Average Number of Steps Taken vs 5-min Interval", y="Average Number of Steps", x="5-min Interval Times Series")
plot.step.interval

optimal <- which.max(activity3$mean.step)
optimal.step <- activity3$interval[optimal]
```

Maximum number of steps is coming from `r optimal.step`th 5-min interval 

## Imputing missing values
```{r,echo=TRUE}
# Total number of missing values in the dataset
sum(is.na(activity))
```

Total number of missing values in the dataset is `r sum(is.na(activity))`.

```{r,echo=TRUE}
impute.activity <- activity
impute.activity$steps[is.na(impute.activity$steps)] <- mean(impute.activity$steps,na.rm=TRUE)
impute.activity$steps <- as.numeric(impute.activity$steps)
impute.activity$interval <- as.numeric(impute.activity$interval)
colSums(is.na(impute.activity))

# Summarize data by date
impute.activity2 <- summarize(group_by(impute.activity,date),daily.step=sum(steps))

mean.impute   <- as.integer(mean(impute.activity2$daily.step))

median.impute <- as.integer(median(impute.activity2$daily.step))

# Plot histogram
plot.steps.day <- ggplot(impute.activity2, aes(x=daily.step)) + 
  geom_histogram(binwidth = 1000, aes(y=..count.., fill=..count..)) + 
  geom_vline(xintercept=mean.impute, colour="red", linetype="dashed", size=1) +
  geom_vline(xintercept=median.impute, colour="green" , linetype="dotted", size=1) +
  labs(title="Histogram of Number of Steps taken each day (impute)", y="Frequency", x="Daily Steps")
plot.steps.day
```

Mean total number of steps taken per day (after impute) is `r mean.impute`.

Median total number of steps taken per day (after impute) is `r median.impute`.

## Are there differences in activity patterns between weekdays and weekends?
```{r,echo=TRUE}
impute.activity$day <- ifelse(weekdays(impute.activity$date) %in% c("Saturday","Sunday"), "weekday", "weekend")
# Preparing data for ggplot
impute.df <- impute.activity %>% group_by(interval,day) %>% summarise(mean.step=mean(steps))

# Plot Average steps across weekday/weekend vs 5-min interval Time Series
plot.weekday.interval <- ggplot(impute.df, aes(x=interval, y=mean.step, color=day)) + 
  facet_grid(day~.) +
  geom_line() + 
  labs(title="Average Number of Steps Taken vs 5-min Interval on Weekday/Weekend", y="Average Number of Steps", x="5-min Interval Times Series")
plot.weekday.interval
```
